Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-22 manuscript.qmd
clarify introduction

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck are part of the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs. Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients. However, the overall prevalence of involvement in a given LNL does not equal a patient's personal risk for occult disease in that LNL, which depends on an individual patient's state of tumor progression. Currently, a patient presenting with no clinically detectable nodal disease (cN0) and a small, clearly lateralized T1 tumor, the same recommended contralateral CTV-N as a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane. Both patients receive extensive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].

To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019]. So far, these models were limited to the ipsilateral side of the neck. The main contribution of this work is the extension of the formalism to include risk predictions for the contralateral side as well. The model may then guide the reduction of the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low. In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.

@@

1. In section @sec-data we present a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients. Based on the dataset the main risk factors for contralateral lymph node involvement are identified and the requirements for the bilateral model extension are described (@sec-requirements).

2. In @sec-ext-to-contra we present a bilateral HMM of lymphatic progression that accounts for the T-stage lateralization of the primary tumor as well as clinical involvement as risk factors for occult contralateral involvement. In @sec-methods we describe how the model training and the computational experiments were set up.

3. In the results in @sec-results, we demonstrate that the model is able to describe the patterns of contralateral lymph node involvement observed in the dataset, and we apply the model to estimating risk of occult disease for typical patients. In @sec-discussion, implications for volume-deescalated radiotherapy are discussed.

>>>>

In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted. Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels (LNLs) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases within these levels. They often recommend extensive irradiation of both sides of the neck. However, the general prevalence of metastasis in a given LNL does not correspond to an individual patient's risk of occult disease in that region, which depends on their specific state of tumor progression. For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane: Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].

To better quantify individualized risk of occult disease, we previously developed an intuitive probabilistic hidden Markov model (HMM) [@ludwig_hidden_2021;@ludwig_modelling_2023], originally based on a conceptually similar a Bayesian network model [@pouymayou_bayesian_2019]. However, these models have been limited to predicting ipsilateral nodal involvement. This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations. By identifying patients with low contralateral risk, the model could guide reductions in contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.

@@

1. Section @sec-data presents a multi-centric dataset on lymph node involvement in {{< var data.num_patients >}} OPSCC patients, identifying key risk factors for contralateral lymph node involvement and outlining requirements for a bilateral model extension (@sec-requirements).

2. Section @sec-ext-to-contra introduces a bilateral HMM that incorporates primary tumor lateralization, T-category, and clinical involvement as risk factors for contralateral involvement. Model training and computational experiments are described in @sec-methods.

3. Section @sec-results demonstrates the model's ability to replicate observed contralateral lymph node involvement patterns and estimates occult disease risk for typical patients. Implications for volume-deescalated radiotherapy are discussed in @sec-discussion.
